---
title: "Untitled"
format:
  pdf:
    documentclass: apa7
editor: visual
keep-tex: true
---

<!-- \usepackage{lipsum} -->

<!-- \usepackage[american]{babel} -->

<!-- \usepackage{csquotes} -->

<!-- \usepackage[style=apa,sortcites=true,sorting=nyt,backend=biber]{biblatex} -->

<!-- \DeclareLanguageMapping{american}{american-apa} -->

<!-- \addbibresource{bibliography.bib} -->

<!-- \title{Network visualization in R using the netplot package} -->

<!-- \shorttitle{Sample Document} -->

<!-- \authorsnames[{1,2},2]{Porter Bischoff, George G. {Vega Yon}} -->

<!-- \authorsaffiliations{Utah Valley University, University of Utah} -->

<!-- \leftheader{Weiss} -->

<!-- \abstract{\lipsum[1]} -->

<!-- \keywords{APA style, demonstration} -->

<!-- \authornote{ -->

<!--    \addORCIDlink{Daniel A. Weiss}{0000-0000-0000-0000} -->

<!--   Correspondence concerning this article should be addressed to Daniel A. Weiss, Department of Educational Psychology, Counseling and -->

<!--   Special Education, A University Somewhere, 123 Main St., Oneonta, NY -->

<!--   13820.  E-mail: daniel.weiss.led@gmail.com} -->

## Introduction

Background, what's out there (visualization tools,) why this is useful (because there are not that many detailed examples showing the code, talk about your experience in Sunbelt "what's the format of the data", look for papers talking about computing literacy) and our goal (start to finish network visualization: load the data, process it a little bit, and plot it).

## Network visualization in a nutshell: Things to consider

Talk about the different aspects about network viz the user needs to consider: layout, vertex size, vertex colour, vertex shape, edges, edges width, etc. Talk about the different components and how can we use them (to represent what, for example.) The size of the network, and type of the network (egocentric, small, large, bipartite, etc.)

In terms of the layouts, what are the things we need to consider (we can mention R packages that implement layouts in R).

## Start to finish example

### Example 1: School data

#### Cleaning Data 

First, the data needs to be pulled in. It can be found at (INSERT LINK HERE). After we pull it in, let's get a glimpse of what the data looks like.

```{r}
#| echo: true
#| include: true
#| label: packages
#| warning: false
#| message: false

# attaching packages
library(igraph)
library(data.table)
library(devtools)
install_github("USCCANA/netplot")
library(netplot)


# loading and cleaning data
students <- fread("./data/middle_school/pone.0153690.s001.csv")
interactions <- fread("./data/middle_school/pone.0153690.s003.csv")

print(students)
print(interactions)
```


In order to use the data, we need to remove all of the 'N/A's and miscoding in the datasets. Also, we see a large number of students who only have interactions with themselves (they do not interact with anyone else through the day), so these "isolates" need to be removed in order for the graph to be more easily read. 
```{r}
#| echo: true
#| include: true
#| label: filter
#| warning: false
#| message: false

# filtering out 'N/A's in the 'students' data frame
students <- students[!is.na(id)]

# filtering down to gender being "0" or "1"
students <- students[gender %in% c("0", "1")]

# filter out 'N/A's in 'id' and 'contactId' 
interactions <- interactions[!is.na(id) & !is.na(contactId)]

# Which connections are not OK?
ids <- sort(unique(students$id))

# narrowed our data from 10781 to 5150
interactions <- interactions[(id %in% ids) & (contactId %in% ids)]

source(file = "./misc/color_nodes_function.R")
```


After, the two datasets need to be combined together. 
```{r}
#| echo: true
#| include: true
#| label: combine
#| warning: false
#| message: false

## Creating matrix from datasets
net <- graph_from_data_frame(
  d = interactions[, .(id, contactId)],
  directed = FALSE, vertices = as.data.frame(students)
)

## Getting only connected individuals
net_with_no_isolates <- induced_subgraph(net, which(degree(net) > 0))
```


Finally, we plot it, effectively showing this network graph. 
```{r}
#| echo: true
#| include: true
#| label: plot_school1
#| warning: false
#| message: false

## Plot with no isolates
nplot(
  net_with_no_isolates
) 
```


#### Vertex Options 
Here, we are taking the data set and the plot, letting us customize a number of aspects of the graph. First, in order to work with the "color_nodes" function, we need to make "grade" a factor instead of being numeric. Also, we identify the colors we would like the nodes to be.
```{r}
#| echo: true
#| include: true
#| label: grade
#| warning: false
#| message: false

  ## adjust 'grade' to factor 
V(net_with_no_isolates)$grade <-  as.factor(V(net_with_no_isolates)$grade)  

# plotting connections among grades ####
set.seed(77) 

a_colors <- color_nodes(net_with_no_isolates,"grade", c("gray40","red3"))
attr(a_colors, "map")
```

Now, we are able to create a plot of the data. This is the same data that we used to create the plot above, but now adjustments to the nodes will be made. 

- Color the vertices ('vertex.color') according to the grade the student is in (with 7th graders being gray and 8th graders being red).

- Adjust the shape of the vertices ('vertex.nsides'). If the student is a 7th grader, the vertices will be a circle, but if they are not, the vertices will be a triangle.

- Adjust size of vertices ('vertex.size.range'). 

- Remove the labels of the nodes. 

```{r}
#| echo: true
#| include: true
#| label: grade_vertices
#| warning: false
#| message: false

grades <- nplot(
  net_with_no_isolates,
  vertex.color = color_nodes(net_with_no_isolates, "grade", c("gray40","red3")),
  vertex.nsides = ifelse(V(net_with_no_isolates)$grade == 7, 10, 3),
  vertex.size.range = c(0.015, 0.020),
  vertex.label=NULL)

print(grades)
```

This looks good, but lets alternate these parameters we just gave to make things have a different look.

- Change `vertex.colors` to be tied to a color palette. 

- Adjust `vertex.nsides` to make 7th graders be an octagon and 8th graders be a hexagon. 

- Adjust `vertex.size.range`, making each vertex smaller.

- Add and adjust labels of vertices with functions `vertex.label.[specific_function]`
  
  - `vertex.label.fontsize` adjust the font size
  
  - `vertex.label.show` adjusts proportion of labels to keep. 

- Adjust `vertex.frame.color` to give an outline of each vertex.

- Adjust `bg.col` to give a background color to the graph.

```{r}
#| echo: true
#| include: true
#| label: grade_vertices1
#| warning: false
#| message: false

library(igraph)
library(RColorBrewer)

# Create a color palette using RColorBrewer
palette <- brewer.pal(3, "Set1")  # Change the number and palette name as needed

grades <- nplot(
  net_with_no_isolates,
  vertex.color           = color_nodes(net_with_no_isolates, "grade", palette),
  vertex.nsides          = ifelse(V(net_with_no_isolates)$grade == 7, 8, 6),
  vertex.size.range      = c(0.01, 0.011),
  vertex.label.fontsize  = 10,
  vertex.label.show      = .25,
  vertex.frame.color     = "black",
  bg.col                 = "mistyrose")

print(grades)

```

#### Edge Options
Now that we have explored a bit about vertices, let's dive into options related to edges. 

- Change `edge.width.range` to make the size of the edges wider or thinner. 

- Change `edge.arrow.size` to add arrows to the graph. 
```{r}
grades <- nplot(
  net_with_no_isolates,
  edge.width.range = c(.9,1))

print(grades)
```

